Minimax-optimal learning with graph Laplacians

### 2020-08-05 notes/graph_regression2.tex
Added theorem on estimation rates for Laplacian smoothing.

In graph based learning, we perform the above inferential tasks by first building a neighborhood graph $G_{n,r}$, which should approximately capture the geometry of $P$ and $\mc{X}$, and then constructing an estimate $\wh{f}$ which is smooth with respect to $G_{n,r}$. For a kernel function $K: [0,\infty) \to \Reals$ and connectivity radius $r > 0$, the neighborhood graph $G_{n,r} = ([n],{\bf W})$ is a weighted, undirected graph on vertices $[n] = \{1,...,n\}$, which we associate with the samples $\{X_1,\ldots,X_n\}$. The weight matrix ${\bf W} = ({\bf W}_{ij})_{ij}$ is an $n \times n$ matrix with entries
\begin{equation*}
\label{eqn:neighborhood_graph}
{\bf W}_{ij} = K\Biggl(\frac{\norm{X_i - X_j}_{\Rd}}{r}\Biggr),
\end{equation*}
and the degree matrix ${\bf D}$ is the $n \times n$ diagonal matrix with entries ${\bf D}_{ii} = \sum_{j = 1}^{n}{\bf W}_{ij}$.  Our star operator, the neighborhood graph Laplacian, can then be written as
\begin{equation}
\label{eqn:graph_Laplacian}
\Lap_{n,r} = \bf{D} - \bf{W}
\end{equation}
\emph{Laplacian eigenmaps} and \emph{Laplacian smoothing} each use $\Lap_{n,r}$ to construct an estimator $\wh{f}$ of $f_0$. Letting $v_1(G_{n,r}),\ldots,v_n(G_{n,r})$ denote eigenvectors of $\Lap_{n,r}$, the Laplacian eigenmaps estimator $\wh{f}_{\LE}$ \citep{belkin2003} is given by 
\begin{equation}
\label{eqn:laplacian_eigenmaps}
\wh{f}_{\textrm{LE}} = \sum_{k = 1}^{\kappa} \Dotp{Y}{v_k(G_{n,r})}_n v_{k}(G_{n,r})
\end{equation}
where $\kappa \in \{1,...,n\}$ is a tuning parameter. (Here $\dotp{u}{v}_n = \frac{1}{n}\sum_{i = 1}^{n} u_i v_i$ for vectors $u,v \in \Reals^n$.) The Laplacian smoothing estimator $\wh{f}_{\LS}$ \citep{smola2003} is given by
\begin{equation}
\label{eqn:laplacian_smoothing}
\wh{f}_{\LS}= \min_{f \in \Reals^n} \Bigl\{\sum_{i = 1}^{n}(Y_i - f_i)^2 + \rho \cdot f^T L^s f \Bigr\}
\end{equation}
where $\rho > 0$ acts a tuning parameter. Then, assuming~\eqref{eqn:laplacian_eigenmaps} and~\eqref{eqn:laplacian_smoothing} are reasonable estimators of $f_0$, the size of the statistics
\begin{align}
T_{\LE} & = \norm{\wh{f}_{\LE}}_n^2 \label{eqn:laplacian_eigenmaps_test}\\
T_{\LS} & = \norm{\wh{f}_{\LS}}_n^2 \label{eqn:laplacian_smoothing_test}
\end{align}
can be used to test whether or not $f_0 = 0$.

@@

	\begin{enumerate}[(P1)]
		\item 
		\label{asmp:bounded_density}
		The distribution $P$ is defined on domain $\Xset = [0,1]^d$, and admits a density $p$ with respect to the uniform (Lebesgue) measure on $[0,1]^d$. The density $p$ is bounded away from $0$ and $\infty$,
		\begin{equation*}
		0 < p_{\min} < p(x) < p_{\max} < \infty
		\end{equation*}
		for all $x \in [0,1]^d$.
	\end{enumerate}
	The assumption~\ref{asmp:bounded_density} is a standard but significant assumption; we will postpone a more significant discussion of it until \textcolor{red}{a later point}. Assuming~\ref{asmp:bounded_density}, the minimax estimation rate over the Sobolev ellipsoids $H^{s}(P,M)$ is
	\begin{equation*}
	\inf_{\wh{f}} \sup_{f \in H^s(P,M)} \Ebb\Bigl[\norm{\wh{f} - f}_{L^2(P)}^2\Bigr] \asymp n^{-2s/(2s + d)}~~\textrm{for all $s$ and $d$,}
	\end{equation*}
	where the infimum is over all measurable maps $\wh{f}(X_1,Y_1,\ldots,X_n,Y_n)$
	\item \textbf{Minimax hypothesis testing.}
	\item \textbf{Our contributions, long version.}
	\begin{itemize}
		\item \textbf{Minimax optimal estimation.}
		We show that with high probability,  $\norm{\wh{f}_{\LE} - f_0}_{n}^2 \lesssim n^{-2s/(2s + d)}$ whenever $f_0 \in C^s(\mc{X})$ for any $s \leq 1$, or $f_0 \in H^s(P)$ for $s = 1$. The same conclusions hold with respect to $\wh{f}_{\LS}$ when $d \leq 4$. 
		\item \textbf{Minimax optimal testing.}
		We show that with high probability, a level-$\alpha$ test constructed using $T_{\LE}$ has non-trivial power whenever $f_0 \in C^s(\mc{X})$ for any $s \leq 1$---or $f_0 \in H^s(P)$ for $s = 1$---additionally satisfies $\norm{f_0}_{\Leb^2(P)} \gtrsim \max\{n^{-4s/(4s + d)},n^{-1/4}\}$. \textcolor{red}{The same conclusions hold with respect to $\wh{f}_{LS}$.}
		\item \textbf{Manifold adaptivity.}
		If $\mc{X} \subset \Rd$ is a submanifold of dimension $m < d$, each of the aforementioned rates hold with $d$ replaced by $m$.
		\item \textbf{Taking advantage of higher-order smoothness conditions.} Under appropriate boundary conditions on the function classes $C^s$ and $H^s$ and additional smoothness conditions on $P$, when $s > 1$  each of the aforementioned conclusions hold for certain values of $d$.
	\end{itemize}
	\item \textbf{Organization.}
	\item \textbf{Notation.}
\end{itemize}

@@

The estimators~\eqref{eqn:laplacian_eigenmaps} and~\eqref{eqn:laplacian_smoothing}, and the resulting test statistics~\eqref{eqn:laplacian_eigenmaps_test} and ~\eqref{eqn:laplacian_smoothing_test}, achieve the standard minimax optimal estimation and hypothesis testing rates, respectively.

@@

Under appropriate conditions, the Laplacian eigenmaps estimator $\wh{f}_{\LE}$ achieves minimax rates over all $f_0 \in H_0^s(P,M)$ with high probability. These conditions will take two forms. The first pair of conditions will be on the neighborhood graph $G_{n,r}$. These assumptions are mild, as the construction of the graph is in the user's control.
\begin{enumerate}[label=(K\arabic*)]
	\item 
	\label{asmp:kernel_form}
	The kernel function $K(z) = \1\{\norm{z}_2/r \leq 1 \}$.
	\item 
	\label{asmp:kernel_radius}
	The radii $r = r(n)$ satisfy $r(n) \to 0$ as $n \to \infty$, and additionally
	\begin{equation*}
	n^{-1/(2(s - 1) + d)}\leq r(n) \leq M^{2/(2s + d)}n^{-1/(2s + d)}
	\end{equation*}
	for all $n$ sufficiently large.
\end{enumerate}

@@

\section{Higher-order adaptivity}

>>>>

In graph based learning, we perform the above inferential tasks by first building a neighborhood graph $G_{n,r}$, which should approximately capture the geometry of $P$ and $\mc{X}$, and then constructing an estimate $\wh{f}$ which is smooth with respect to $G_{n,r}$. For a kernel function $K: [0,\infty) \to \Reals$ and connectivity radius $r > 0$, the neighborhood graph $G_{n,r} = ([n],{\bf W})$ is a weighted, undirected graph on vertices $[n] = \{1,...,n\}$, which we associate with the samples $\{X_1,\ldots,X_n\}$. The weight matrix ${\bf W} = ({\bf W}_{ij})_{ij}$ is an $n \times n$ matrix with entries
\begin{equation*}
\label{eqn:neighborhood_graph}
{\bf W}_{ij} = K\Biggl(\frac{\norm{X_i - X_j}_{\Rd}}{r}\Biggr),
\end{equation*}
and the degree matrix ${\bf D}$ is the $n \times n$ diagonal matrix with entries ${\bf D}_{ii} = \sum_{j = 1}^{n}{\bf W}_{ij}$.  Our star operator, the neighborhood graph Laplacian, can then be written as
\begin{equation}
\label{eqn:graph_Laplacian}
\Lap_{n,r} = \bf{D} - \bf{W}
\end{equation}
\emph{Laplacian eigenmaps} and \emph{Laplacian smoothing} each use $\Lap_{n,r}$ to construct an estimator $\wh{f}$ of $f_0$. Letting $v_1(G_{n,r}),\ldots,v_n(G_{n,r})$ denote eigenvectors of $\Lap_{n,r}$, the Laplacian eigenmaps estimator $\wh{f}_{\LE}$ \citep{belkin2003} is given by 
\begin{equation}
\label{eqn:laplacian_eigenmaps}
\wh{f}_{\textrm{LE}} = \sum_{k = 1}^{\kappa} \Dotp{Y}{v_k(G_{n,r})}_n v_{k}(G_{n,r})
\end{equation}
where $\kappa \in \{1,...,n\}$ is a tuning parameter. (Here $\dotp{u}{v}_n = \frac{1}{n}\sum_{i = 1}^{n} u_i v_i$ for vectors $u,v \in \Reals^n$.) The Laplacian smoothing estimator $\wt{f}_{\LS}$ \citep{smola2003} is given by
\begin{equation}
\label{eqn:laplacian_smoothing}
\wt{f}_{\LS}= \min_{f \in \Reals^n} \Bigl\{\sum_{i = 1}^{n}(Y_i - f_i)^2 + \rho \cdot f^T L^s f \Bigr\}
\end{equation}
where $\rho > 0$ acts a tuning parameter. Then, assuming~\eqref{eqn:laplacian_eigenmaps} and~\eqref{eqn:laplacian_smoothing} are reasonable estimators of $f_0$, the size of the statistics
\begin{align}
T_{\LE} & = \norm{\wh{f}_{\LE}}_n^2 \label{eqn:laplacian_eigenmaps_test}\\
T_{\LS} & = \norm{\wt{f}_{\LS}}_n^2 \label{eqn:laplacian_smoothing_test}
\end{align}
can be used to test whether or not $f_0 = 0$.

@@

	\begin{enumerate}[(P1)]
		\item 
		\label{asmp:bounded_density}
		The distribution $P$ is defined on domain $\Xset = [0,1]^d$, and admits a density $p$ with respect to the uniform (Lebesgue) measure on $[0,1]^d$. The density $p$ is bounded away from $0$ and $\infty$,
		\begin{equation*}
		0 < p_{\min} < p(x) < p_{\max} < \infty
		\end{equation*}
		for all $x \in [0,1]^d$.
	\end{enumerate}
	The assumption~\ref{asmp:bounded_density} is a standard but significant assumption; we will postpone a more significant discussion of it until \textcolor{red}{a later point}. Assuming~\ref{asmp:bounded_density}, the minimax estimation rate over the Sobolev ellipsoids $H^{s}(P,M)$ is
	\begin{equation*}
	\inf_{\wh{f}} \sup_{f \in H^s(P,M)} \Ebb\Bigl[\norm{\wh{f} - f}_{L^2(P)}^2\Bigr] \asymp n^{-2s/(2s + d)}~~\textrm{for all $s$ and $d$,}
	\end{equation*}
	where the infimum is over all measurable maps $\wh{f}(X_1,Y_1,\ldots,X_n,Y_n)$
	\item \textbf{Minimax hypothesis testing.}
	\item \textbf{Our contributions, long version.}
	\begin{itemize}
		\item \textbf{Minimax optimal estimation.}
		We show that with high probability,  $\norm{\wh{f}_{\LE} - f_0}_{n}^2 \lesssim n^{-2s/(2s + d)}$ whenever $f_0 \in C^s(\mc{X})$ for any $s \leq 1$, or $f_0 \in H^s(P)$ for $s = 1$. The same conclusions hold with respect to $\wt{f}_{\LS}$ when $d \leq 4$. 
		\item \textbf{Minimax optimal testing.}
		We show that with high probability, a level-$\alpha$ test constructed using $T_{\LE}$ has non-trivial power whenever $f_0 \in C^s(\mc{X})$ for any $s \leq 1$---or $f_0 \in H^s(P)$ for $s = 1$---additionally satisfies $\norm{f_0}_{\Leb^2(P)} \gtrsim \max\{n^{-4s/(4s + d)},n^{-1/4}\}$. \textcolor{red}{The same conclusions hold with respect to $\wh{f}_{LS}$.}
		\item \textbf{Manifold adaptivity.}
		If $\mc{X} \subset \Rd$ is a submanifold of dimension $m < d$, each of the aforementioned rates hold with $d$ replaced by $m$.
		\item \textbf{Taking advantage of higher-order smoothness conditions.} Under appropriate boundary conditions on the function classes $C^s$ and $H^s$ and additional smoothness conditions on $P$, when $s > 1$  each of the aforementioned conclusions hold for certain values of $d$.
	\end{itemize}
	\item \textbf{Organization.}
	\item \textbf{Notation.}
\end{itemize}

@@

Let us start by assuming $f_0 \in H^1(P,M)$. In this case, the estimators~\eqref{eqn:laplacian_eigenmaps} and~\eqref{eqn:laplacian_smoothing}, and the resulting test statistics~\eqref{eqn:laplacian_eigenmaps_test} and~\eqref{eqn:laplacian_smoothing_test}, achieve the standard minimax optimal estimation and hypothesis testing rates, respectively.

@@

Under appropriate conditions, the Laplacian eigenmaps estimator $\wh{f}_{\LE}$ achieves minimax rates over all $f_0 \in H^1(P,M)$ with high probability. The conditions we impose will be on the neighborhood graph $G_{n,r}$. As the construction of the graph is in the user's control, these assumptions are therefore mild.
\begin{enumerate}[label=(K\arabic*)]
	\item 
	\label{asmp:kernel_form}
	The kernel function $K(z) = \1\Bigl\{\norm{z}_2/r \leq 1 \Bigr\}$.
	\item 
	\label{asmp:kernel_radius}
	The radii $r = r(n)$ satisfies $r(n) \to 0$ as $n \to \infty$, and additionally
	\begin{equation*}
	n^{-1/d}\leq r(n) \leq M^{2/(2 + d)}n^{-1/(2 + d)}
	\end{equation*}
	for all $n$ sufficiently large.
\end{enumerate}

We use~\ref{asmp:kernel_form} and~\ref{asmp:kernel_radius} to ensure that the approximation error (bias) of $\wh{f}_{\LE}$ is of the correct order. On the other hand, bounding the estimation error of $\wh{f}_{\LE}$ -- both in expectation and in probability-- is straightforward. Together, bounds on these two sources of error imply Theorem~\ref{thm:laplacian_eigenmaps_estimation1}.
\begin{theorem}
	\label{thm:laplacian_eigenmaps_estimation1}
	Let $(X_1,Y_1),\ldots,(X_n,Y_n)$ be random samples drawn according to the model~\eqref{eqn:random_design_regression}. Suppose that $f_0 \in H^1(P,M)$, that $P$ satisfies~\ref{asmp:bounded_density}, and that neighborhood graph $G_{n,r}$ is computed with kernel $K$ and radius $r$ which satisfy~\ref{asmp:kernel_form} and~\ref{asmp:kernel_radius}. Then, for the Laplacian eigenmaps estimator $\wh{f}_{\LE}$ as given by~\eqref{eqn:laplacian_eigenmaps}, when $\kappa = M^{2/(2 + d)}n^{d/(2 + d)}$, the following statement holds: for all $n$ sufficiently large, there exist constants $c$ and $C$ such that
	\begin{equation*}
	\Bigl\|\wh{f}_{\LE} - f_0\Bigr\|_n^2 \leq \frac{C}{\delta} M^{2d/(4 + d)} n^{-2/(2 + d)}
	\end{equation*}
	with probability at least $1 - \delta -  Cr^{-d}\exp\{-cnr^d\} - \exp(-\kappa)$.
\end{theorem}

\begin{itemize}
	\item Discuss the choice of radius $r$ imposed by~\ref{asmp:kernel_radius}.
	\item Comment on the distinction between error rates as measured by the norm in $L^2(P_n)$ vs. $L^2(P)$. 
\end{itemize}

Theorem~\ref{thm:laplacian_smoothing_estimation1} is analogous to Theorem~\ref{thm:laplacian_eigenmaps_estimation1}, dealing with the Laplacian smoothing estimator $\wt{f}_{\LS}$. The key difference is that it holds only when the dimension $d < 4$.  
\begin{theorem}
	\label{thm:laplacian_smoothing_estimation1}
	Let $(X_1,Y_1),\ldots,(X_n,Y_n)$ be random samples drawn according to the model~\eqref{eqn:random_design_regression}. Suppose that $f_0 \in H^1(P,M)$, that $P$ satisfies~\ref{asmp:bounded_density}, and that neighborhood graph $G_{n,r}$ is computed with kernel $K$ and radius $r$ which satisfy~\ref{asmp:kernel_form} and~\ref{asmp:kernel_radius}. Then, for the Laplacian eigenmaps estimator $\wh{f}_{\LE}$ as given by~\eqref{eqn:laplacian_eigenmaps}, when $\kappa = M^{2/(2 + d)}n^{d/(2 + d)}$, the following statement holds: for all $n$ sufficiently large, there exist constants $c$ and $C$ such that
	\begin{equation*}
	\Bigl\|\wh{f}_{\LE} - f_0\Bigr\|_n^2 \leq \frac{C}{\delta} M^{2d/(4 + d)} n^{-2/(2 + d)}
	\end{equation*}
	with probability at least $1 - \delta -  Cr^{-d}\exp\{-cnr^d\} - \exp(-\kappa)$.
\end{theorem}

\begin{itemize}
	\item A restriction on the maximum value of dimension $d$ such as $d \leq 4$ is unsurprising. In a related context (estimation over a $d$-dimensional lattice graph), \cite{sadhanala2016} require the same restriction. The underlying phenomenon is similar to that elucidated by the seminal papers~\cite{birge1993} and~\cite{birge1998}: when the dimension $d$ is too large, the entropy function is irregular and (penalized) least squares approaches are no longer minimax optimal.
\end{itemize}

\section{Higher-order smoothness classes.}
\label{sec:higher_order_smoothness_classes}
Suppose we are willing to assume $f_0$ possesses more than one smooth derivative. We can show that in this situation, the estimators~\eqref{eqn:laplacian_eigenmaps} and~\eqref{eqn:laplacian_smoothing}, and test statistics~\eqref{eqn:laplacian_eigenmaps_test} and~\eqref{eqn:laplacian_smoothing_test} obtain the faster (known) minimax rates in the nonparametric estimation and testing problems. However, these conclusions hold only for certain values of the dimension $d$.

\subsection{Minimax optimal estimation}
Under  conditions, the Laplacian eigenmaps estimator $\wh{f}_{\LE}$ achieves minimax rates over all $f_0 \in H_0^s(P,M)$ with high probability. These conditions will take two forms. The first pair of conditions will be on the neighborhood graph $G_{n,r}$. These assumptions are mild, as the construction of the graph is in the user's control.
\begin{enumerate}[label=(K\arabic*)]
	\item 
	\label{asmp:kernel_form}
	The kernel function $K(z) = \1\{\norm{z}_2/r \leq 1 \}$.
	\item 
	\label{asmp:kernel_radius}
	The radii $r = r(n)$ satisfy $r(n) \to 0$ as $n \to \infty$, and additionally
	\begin{equation*}
	n^{-1/(2(s - 1) + d)}\leq r(n) \leq M^{2/(2s + d)}n^{-1/(2s + d)}
	\end{equation*}
	for all $n$ sufficiently large.
\end{enumerate}

@@
